Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors for COVID-19 in Scotland (REACT-SCOT): a population-based case-control study

### 2020-07-12 response_reviewers.Rmd

---
title: "PMEDICINE-D-20-02483R1: authors' response to reviewers' comments"

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
output: pdf_document

# rmarkdown::render("response_reviewers.md")

---

# Requests from the editors:

> Please ensure that the paper adheres to the PLOS Data Availability Policy

We have included the following data availability statement on the submission portal and in the manuscript:

The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data. All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on https://github.com/pmckeigue/covid-scotland_public

> Please structure the abstract using the PLOS Medicine headings (Background, Methods and Findings, Conclusions - "Methods and Findings" should be a single subsection heading).

Done

> In the last sentence of the Abstract Methods and Findings section, please include a brief note about any key limitation(s) of the study's methodology.

We have added a sentence to note that records from primary care were not available.

> At this stage, we ask that you include a short, non-technical Author Summary of your research to make findings accessible to a wide audience that includes both scientists and non-scientists. The Author Summary should immediately follow the Abstract in your revised manuscript. This text is subject to editorial change and should be distinct from the scientific abstract. Please see our author guidelines for more information: https://journals.plos.org/plosmedicine/s/revising-your-manuscript#loc-author-summary

Done

> Did your study have a prospective protocol or analysis plan? Please state this (either way) early in the Methods section.

We have inserted this sentence into the Methods: "The study followed a pre-specified protocol. Modifications after study started were that we aligned the list of conditions of interest to be consistent with those designated as moderate risk conditions by public health agencies, and extended the list of drug classes under study to include all drugs."

> The authors could consider using an appropriate reporting guideline (eg, RECORD - https://www.equator-network.org/reporting-guidelines/record/ - designed to support reporting of observational studies using routinely-collected data) to enhance reporting of methods and findings in their study. If the authors choose to use this please upload the completed RECORD (or other) checklist as supporting information with the revised paper.

RECORD checklist uploaded

# Note for all reviewers. 
As discussed with the editor in this revised manuscript we have used an updated dataset that includes all deaths certified with COVID-19 as underlying cause, irrespective of whether a positive nucleic acid test for SARS-CoV-2 was obtained. The data capture was extended from May 15th to June 8th.

This update has changed all the numbers in the manuscript, but none of the conclusions have changed.  Our reproducible research pipeline allows us to regenerate the entire manuscript from this updated dataset with a single keystroke.

Using this update deals with several points raised by reviewers, including the inclusion of deaths among those who did not test positive, and the incomplete recording of ethnicity.

# Reviewer #1:

> Full case ascertainment of severe disease arguably limited by LTC deaths that do not reach hospital not accounted for, and may bias the rate ratios described.

 As noted above, this updated dataset includes deaths certified with COVID-19 as underlying cause among people without a positive test for SARS-CoV-2.  These are tabulated separately in Supplementary Table 1.

> Challenges are the sheer number of univariate comparisons reported, and the limited  corrections for multiple comparisons in place.

By reporting exact p–values, we allow readers to make their own assessment of the strength of evidence for association allowing for the size of the prior hypothesis space (number of independent tests).  This practice is endorsed by Reviewer #2.

> Table 5 and 6 could be placed in the supplementary appendix, unless the purpose is to emphasize the non-specificity of the associations.

We have moved these tables to the supplementary appendix

> The presence of less robust associations in common conditions such as ischemic heart disease, and more robust associations in rare diseases like chronic kidney disease, is notable. This is mentioned in the discussion, but should be understood better, given the conclusions.

We have added a sentence to note the relevance of non-ischaemic heart disease and kidney disease to the involvement of these systems in severe COVID-19.

> A great deal of the Results should be moved to the Methods, such as the multivariable model construction, and the prescription association data.

We have moved the material describing the stepwise regression procedure to the Methods section

> Removing neoplasms from the prespecified list because of inability to discern who has already been shielded belies the fact that shielding is i) not 100% applied in populations; ii) not practiced in other regions, making the association between neoplasm and outcome an important one to consider for policy in other regions.

We have rewritten this sentence to clarify that neoplasms are not on the designated list of conditions that the National Health Service uses to classify "people at moderate risk", though primary neoplasms of blood or bone marrow and cancer treatments affecting the immune system are on the list used to classify "clinically extremely vulnerable" individuals who are advised to shield themselves.  We have added a sentence to the Discussion to state that we are unable to investigate adequately the risk associated with neoplasms in this study, but that we hope to investigate it in a further study that will include linkage to records of shielding advice.

> Abbreviation 'scrip' in Table 4 not universal, would use 'prescription'

We have made this substitution.

> Very hesitant, given the sample size, the inaccurate labelling problem, the exclusive examination of one ethnic group, and the limited generalizability, to include any reporting or conclusions on the ethnicity-based risk factors. Would suggest excluding altogether, and reporting on the need for more robust ethnicity-based data collection strategies

This updated dataset has more complete recording of self assigned ethnicity as the ascertained was widened to include more SMR datasets and longer lookback thus yielding data on 85% of cases and 74% of controls.  Therefore we have dropped the sensitivity analysis based on name classification which lacked accuracy as noted by the referee.  We now describe the numbers and rate ratios for South Asia, Black and other. We agree that the numbers are too small to make definitive conclusions. However if we do not include any data on ethnicity readers will query this given the amount of concern and publicity there has been about large ethnic differences in risk in the rest of the UK.   Therefore we have made a clear statement in discussion that the numbers are too small to provide definitive statements on risk ratios.

# Reviewer #2:

> Page 3 I understand (probably from Sheila Bird) that death registration in Scotland is much more timely than south of the border but is there any important bias due to delays? I might just give a little bit more detail here as death registration practices are not the same all over the world.

We have edited the Methods section to give the cutoff date for ascertainment of death registrations (rather than deaths) and to note that 94% of deaths among cases were registered within 5 days.

> Page 3 It took me a couple of attempts to see what the sentence starting ‘By restricting’ meant. How about ‘Using this definition means we include all severe cases even if they die without entering critical care.’  That might be too abbreviated so feel free to re–write.

We have rewritten this sentence for clarity as suggested

> Page 3 In my idiolect embarkation means boarding a ship. Does this mean emigration? Whatever it means did the authors really start with 2755 strata of size 10 + 1 and end up with them all of 7 + 1 as this seems to imply. I think I have misunderstood here. I also note that (Table 1) 19670 2755 = 7.14 not 7.

We have rewritten the section on selection of matched controls. In this updated dataset the exclusion of those no longer in Scotland was done before selection of matched controls, so we are closer to the original design target of ten controls per case

> Pages 4 and 5 I think it would be good to cite both gam and survival properly. It is the only credit package authors get. (Disclaimer: I am a package authors although not of these.)

We have added the citation given by the authors of the `survival` package.  For the `gam` package, the author does not give a specific citation.

> Page 5 If the incidence and mortality figures are a core part of the authors’ thesis then I would suggest making some comparison between this and the corresponding figures for all–cause mortality. I sense though that this is perhaps just background. Is there any reason to give overall figures and then just men?

Others have compared COVID-19 mortality with all-cause mortality: a full exploration of this important question is beyond the scope of this paper.

Our figure shows mortality and incidence for men and women separately -- it's possible that the lines for women were not visible to the reviewer.

> Page 6 If the authors are going, correctly, to do a sensitivity analysis for different ways of identifying ethnicity then should it not be flagged in the methods? > Page 6 Despite what it says here Table 1 does show the figures for Black as well as White and South Asian and the risk ratio for Black seems indistinguishable from that for South Asian.

We have deleted the sensitivity analyses regarding using name-based ethnicity classification as we now have more complete recording of  self assigned ethnicity  by doing longer and wider lookback in SMR records.  We now report for White, South Asian, Black and other categories.

> Page 6 The authors state here that Table 3 shows listed conditions broken down by age (so that we see that immune conditions were not more common in 75+). I do not see such a breakdown there.

We have corrected this table reference

> Page 7 The phrase ‘most significant association was with diagnoses in ICD chapter 2’ seems to equate the smallness of the p–value with importance since several of the other risk ratios in Table 5 are of comparable size or even greater than that for neoplasms.

We have rewritten this.

> Page 7 Viewing this from the perspective of a future researcher who finds something apparently different in their data I wonder whether the criteria for Table S5 are too restrictive. I can understand excluding rare conditions (the current limit is 50) but if you restrict by p–value people who get a similar risk ratio to the present authors will not be aware of the similarity if the current authors’ p–value was > 0.001.

We have filtered this table by p-value to keep the length of the table within reasonable limits.  We plan to make more detailed summary statistics available elsewhere.

> Page 8 One bit does not seem very much information to me and I find it hard to see how it would work in practice.

We have added a more detailed explanation of this under "Relevance to policy" in the Discussion -- the total information for discrimination is the sum of the information from the logistic regression on age and sex and the information from the conditional logistic regression on risk factors, conditioned on age and sex.

> Page 8 While it is welcome to see the authors challenge the notion that everybody whose family origins are in South Asia is the same I wonder how much sense this is going to make to people outside the UK or South Asia. International readers may not know how and why people from the disparate communities came to the UK nor why this issue seems so salient in the UK discourse about COVID–19.

Given the comments of Reviewer #1 we have shortened this part of the discussion making clearer the background concern from other studies in the UK but keeping our inference about the data in Scotland brief and noting the wide confidence intervals.

> Page 14 I assume the black dots are hidden behind the black curve?  Strictly speaking multivariable is meant throughout I think (Hidalgo and Goodman, 2013). Multivariate means multiple variables on the left hand side which is not, as far as I can see, true of the models used. I know this is very picky.

We have substituted the term multivariable for multivariate.

> The authors comment on page 8 that they have shown that the risk of a severe outcome increases with age. As far as the death component is concerned that was true of course before this particular virus appeared. According to the English Life Tables ELT17 the value of mx for a male my age was 0.025 and for a male ten years older than me 0.075 which is the same sort of increase with age as the authors are reporting for their severe outcome. What we do not know is what the excess risk is over and above the background risk and how that varies with age.

We have added an explanation in the Discussion of how rate ratios can be interpreted as increases in "Covid age".

-----------------------------------------------------------

# Reviewer #3

> It would be informative if the authors could break down the severe cases into those that were due to death within 28 days without admission to critical care units, those that were admitted to critical care units and survived, those admitted to critical care but perished. It would be interesting to analyze whether what proportion of severe cases among care home residents were death within 28 days without admission to critical care units. Providing the severe cases breakdown and a more in-depth discussion will further strengthen the conclusions.

We have added a supplementary table S1 to provide this breakdown of severe cases, and described what it shows in the first paragraph of Results.  As Reviewer #3 suggested, this shows clearly that almost all fatal cases in care home residents did not enter critical care.

> Since the authors concluded that many preexisting conditions were associated with severe COVID-19, may consider adding ANY comorbidity in Table 3.

The "any comorbidity" variable is implicit in the variable "any prescription or diagnosis".  Other than diabetes which was ascertained from a population register, all other comorbidities are based on dispensed prescriptions or hospital diagnoses.

### 2020-07-12 casecontrol.Rmd
updated pharmaco scripts

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true

@@

@@

# Author summary
Most people infected with the SARS-CoV-2 coronavirus do not become seriously ill.  It is The risk of severe or fatal illness is higher in older than in younger people, and is higher in people with conditions such as asthma and diabetes than in people without these conditions.  Using Scotland's capability for linking electronic health records, we report the first systematic study of the relation of severe or fatal COVID-19 to pre-existing health conditions and other risk factors.  We show that the conditions associated with increased risk include not only those already designated by public health agencies -- asthma, diabetes, heart disease, disabling neurological disease, kidney disease -- but many other diagnoses, associated with frailty and poor health.  However the strongest risk factors are age (and residence in a also asthma, the risk of in younger individuals the most frequent risk-a factor is asthma. 
This lays the basis for constructing risk scores based on electronic health records that can be used to advise people at high risk of severe disease to shield themselves until the epidemic is over.

@@

# Methods
The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)). The pharmacoepidemiological analyses will be reported elsewhere.

## Case definition and selection of matched controls
 All individuals testing positive for nucleic acid for SARS-CoV-2 were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database, which captures all virology testing in all NHS laboratories nationally. Linkage to other datasets was carried out using the Community Health Index (CHI) number, a unique identifier used in all care systems in Scotland.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units and has returned a daily census of patients in critical care from the beginning of the COVID-19 epidemic. Death registrations were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause. Using this definition ensures ascertainment of all severe cases even if they die without testing positive or entering critical care, whatever selection policies may have limited entry to critical care.

For each case, the CHI database was used to select up to ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive and resident in Scotland on the day as the first date that the case tested positive.  For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.  To ensure that cases and controls were representative of the same population at risk, the 0.6% of cases who were recorded on the CHI database as no longer alive and resident in Scotland on the day that ECOSS recorded them as testing positive were also excluded.  As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.  
For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.

@@

@@

# Results
## Comparison of fatal and non-fatal cases
Supplementary Table \ref{tab:casegr} shows a breakdown of severe cases by test-positive status, entry to critical care, and fatal versus non-fatal outcome.  Among those who entered critical care, the median age and the proportion of males were higher in fatal than in non-fatal cases.  Among fatal cases who did not enter critical care. the distribution of age and other risk factors was similar in those with and without a positive test result, except that the proportion of care home residents was higher among those without a positive test result.

## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for male and female mortality are almost parallel.  In models that included age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease. The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease. For severe cases as defined in this study, the sex differential is narrow up to about age 50 but widens between ages 50 and 70 years.  Thus at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but that at later ages the ratio of critical admissions to total fatalities is higher in men.

## Risk factors
### Sociodemographic factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Residence in a care home was by far the strongest risk factor for severe disease.  Higher risk of severe disease was also associated with socioeconomic deprivation.  For the subset of cases and controls in whom ethnicity had been recorded in the Scottish Morbidity Record, the rate ratio (95% CI) associated with South Asian ethnicity with White as reference category was `r table.demog.append["South Asian1", 4]` based on the subset with SMR records. There were vert few cases in other non-White ethnic groups.

@@

@@

Although the assignment of ethnicity is incomplete, the results give some indication of the likely upper bound of the absolute numbers of severe cases in non-White minority ethnic groups in Scotland, and do not show clear evidence of elevated risk in those who identify as South Asian or Black compared to those who identify as White.  This contrasts with reports from England [@the_opensafely_collaborative_opensafely_2020] that have found elevated risks in   non-White groups. In the OpenSAFELY study risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, persisted after adjustment for comorbidities and socioeconomic status.  In a study of risk factors for hospitalized disease in the UK Biobank cohort, adjustment for health care worker status and other social variables attenuated but did not fully explain the elevated crude risk ratios associated with non-White ethnicity [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].  The relative socioeconomic position of ethnic groups in Scotland is different to that in England, so it is plausible that the relation of health status to ethnicity may also differ.  For example in the 2011 Scottish Census 1.6% of the population reported South Asian ethnicity.  Among the 1.0% who identified as Pakistani or Bangladeshi the proportion living in the most deprived neighbourhoods was not higher than the national average [@walsh_increasingly_2019].  Future work may allow more complete assignment of ethnicity and disaggregation of broad categories based on continent of origin.

### Co-morbidities
We have confirmed that the moderate risk conditions designated by the NHS and other agencies [@nhs_whos_2020] are associated with increased risk of severe COVID-19.  However the rate ratios associated with these conditions vary with age - for example the rate ratio associated with diabetes is higher at younger ages. The rate ratios of `r round(as.numeric(substr(table.agegr.all["Type 1 diabetes", 3], 1, 4)), 1)` for Type 1 diabetes and `r round(as.numeric(substr(table.agegr.all["Type 2 diabetes", 3], 1, 4)), 1)` for Type 2 diabetes are broadly similar to those reported in UK Biobank and in the OpenSAFELY studies.  We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.  An unexpected finding was that the risk associated with other forms of heart disease is higher than that associated with ischaemic heart disease.  This category includes conditions such as atrial fibrillation, cardiomyopathies and heart failure. One of the highest rate ratios is that associated with chronic kidney disease.  Prevention of nosocomial transmission in dialysis units may help to reduce this risk.  Over all age groups, `r gsub("(.*\\()(.*)\\)", "\\2", freqs.all["listed.any", 2])` of severe cases had at least one of these listed conditions.  In this dataset it is not possible to examine adequately the risk associated with neoplasms, as we cannot separately identify those who were advised to shield themselves because they had active neoplasms of lymphoid or hematopoietic tissue or were receiving treatments that affect the immune system. We hope to explore this in a separate study that will include linkage to records of shielding advice.  In patients without any listed conditions, further systematic evaluation of past hospitalisation history did not reveal a sparse set of underlying conditions; instead many diagnoses were associated with severe COVID-19.

@@

An important finding of this study was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.  Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that prescription of drugs for the gastrointestinal and central nervous systems, together with nutritional supplements, contributed to this association.  Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, or to more diffuse frailty especially in older persons, causal effects of drugs or direct effects of polypharmacy on susceptibility cannot be ruled out.  These associations are explored in a separate paper.

## Relevance to policy
As lockdown restrictions are eased, there is general agreement that vulnerable individuals will require shielding, even if the restart of the epidemic can be slowed or suppressed by mass testing, contact tracing and isolation of those who test positive.  The "stratify and shield" policy option [@mckeigue_evaluation_2020], in which high-risk individuals comprising up to 15% of the population are shielded for a defined period while the epidemic is allowed to run relatively quickly in low-risk individuals until population-level immunity is attained, depends critically on informative risk discrimination.  So too does the similarly named "segment and shield " option [@bunnik_segmentation_2020] which has the opposite objective of keeping transmissions low.  Although for this preliminary study we have not used the full repertoire of machine learning methods available for this type of problem, we have shown that a model based on health records provides `r table.densities[3, 5]` bits of information for discrimination conditional on age and sex.  Adding this to the `r round(cases.Lambda.agesex, 2)` bits provided by age and sex gives a total information for discrimination of `r round(table.densities[3, 5] + cases.Lambda.agesex, 1)` bits.  We have shown elsewhere that this level of predictive performance would allow at least 80% of those at risk of severe or fatal disease to be allocated to a shielded group comprising no more than 15% of the population [@mckeigue_evaluation_2020]

As awareness grows of how risk varies between individuals, individuals will seek information about their own level of risk.  A key implication of our results is that risk of severe or fatal disease is multifactorial.  The rate ratio of `r round(exp(10 * logistic.coeffs[3, 1]), 1)` associated with a 10-year increase in age is stronger than the rate ratios associated with common diseases such as asthma or Type 2 diabetes that are listed as conditions associated with high risk.  A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one quarter of those aged 60-74 years in the population have at least one of the listed conditions we examined.  It will also exclude many people at high risk because they have multiple risk factors each of small effect.  A more meaningful way to score risk for an individual would be to use all available information to calculate a "COVID age" as the age at which the average risk for someone of the same sex in the population equates to the risk for the individual under study. Thus the rate ratio of `r rr <- round(as.numeric(gsub(" .+", "", table.agegr.all["Type 1 diabetes", 3])), 1); rr` associated with Type 1 diabetes equates to an increase of `r round(log(rr) / logistic.coeffs[3, 1], 1)` years in COVID age.  In Scotland it is technically possible to use existing electronic health records to calculate a risk score for every individual in the population, though more work would be required to develop this as a basis for official advice and individual decisions.

@@

## Protocol registration
The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).

## Data availability
All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).  Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.

@@

## Supplementary material
The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available with this manuscript.

@@

plotWdists(densities=full.densities, distlabels=c("Crude", "Adjusted")) +
                                         theme_grey(base_size = 18) +
                                         theme(legend.title=element_blank())

>>>>

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{soul}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true

@@

delay <- with(cc.severe, table(DATE_OF_REGISTRATION - Date.Death))
percent.withindays <- round(100 * cumsum(delay) / sum(delay))

@@

# Author summary
Most people infected with the SARS-CoV-2 coronavirus do not become seriously ill.  It is The risk of severe or fatal illness is higher in older than in younger people, and is higher in people with conditions such as asthma and diabetes than in people without these conditions.  Using Scotland's capability for linking electronic health records, we report the first systematic study of the relation of severe or fatal COVID-19 to pre-existing health conditions and other risk factors.  We show that the strongest risk factor, apart from age, is residence in a care home. The conditions associated with increased risk include not only those already designated by public health agencies -- asthma, diabetes, heart disease, disabling neurological disease, kidney disease -- but many other diagnoses, associated with frailty and poor health.  This lays a basis for constructing risk scores based on electronic health records that can be used to advise people at high risk of severe disease to shield themselves when there cases in their neighbourhood.

@@

# Methods
\hl{The study followed a pre-specified protocol. Modifications after study started were that we aligned the list of conditions of interest to be consistent with those designated as moderate risk conditions by public health agencies, and extended the list of drug classes under study to include all drugs.}

## Case definition and selection of matched controls
 All individuals testing positive for nucleic acid for SARS-CoV-2 were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database, which captures all virology testing in all NHS laboratories nationally. Linkage to other datasets was carried out using the Community Health Index (CHI) number, a unique identifier used in all care systems in Scotland.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units and has returned a daily census of patients in critical care from the beginning of the COVID-19 epidemic. Death registrations were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause. \hl{Using this definition ensures ascertainment of all severe cases even if they die without testing positive or entering critical care, whatever selection policies may have limited entry to critical care.}

For each case, the CHI database was used to select up to ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive and resident in Scotland on the day as the first date that the case tested positive.  For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.  To ensure that cases and controls were representative of the same population at risk, the 0.6% of cases who were recorded on the CHI database as no longer alive and resident in Scotland on the day that ECOSS recorded them as testing positive were also excluded.  As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.  
For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths registered up to `r format(max(cc.severe$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.  Among fatal cases, `r percent.withindays[6]`% of deaths were registered within 5 days.

@@

Identifiable data were extracted by the Public Health Scotland CHI database and linkage team and de-identified before provision to the analysis team.

@@

# Results
## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for male and female mortality are almost parallel.  In models that included age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease. The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease. For severe cases as defined in this study, the sex differential is narrow up to about age 50 but widens between ages 50 and 70 years.  Thus at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but that at later ages the ratio of critical admissions to total fatalities is higher in men.

## Risk factors
### Sociodemographic factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Residence in a care home was by far the strongest risk factor for severe disease.  Higher risk of severe disease was also associated with socioeconomic deprivation.  In the   `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 2])` of cases  and  `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 1])` of controls  in whom ethnicity had been recorded in the Scottish Morbidity Record, there were few non-White cases or controls and the confidence limits for the risk ratios for severe or fatal disease by ethnic group were wide.

@@

## Comparison of fatal and non-fatal cases

Supplementary Table \ref{tab:casegr} shows a breakdown of severe cases by test-positive status, entry to critical care, and fatal versus non-fatal outcome.   \hl{Severe cases who entered critical care were much younger than severe cases never entering critical care.  The majority of severe cases from residential care homes never entered critical care. Among fatal cases who did not enter critical care. the distribution of age and other risk factors was similar in those with and without a positive test result, except that the proportion of care home residents was higher among those without a positive test result.   Among  those entering critical care  the median age and the proportion of males  and prevalence or prior comorbidities were higher in fatal than in non-fatal cases. }

@@

Although the assignment of ethnicity is incomplete and the prevalence of non-White ethnic groups in Scotland is low, we show the data as they give some indication of the likely upper bound of the absolute numbers of severe cases in South Asian and Black ethnic groups in Scotland. This is relevant because substantial elevations in risk of hospitalised and fatal COVID-19 have been reported elsewhere in the UK [@the_opensafely_collaborative_opensafely_2020; @niedzwiedz_ethnic_2020; @ho_modifiable_2020]. In the OpenSAFELY study risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, persisted after adjustment for comorbidities and socioeconomic status. In Scotland however, although point estimates for rate ratios in those of South Asian and Black ethnicity compared to Whites are above 1, the confidence intervals are too wide for any definitive conclusions on ethnic differences.

### Co-morbidities
We have confirmed that the moderate risk conditions designated by the NHS and other agencies [@nhs_whos_2020] are associated with increased risk of severe COVID-19.  However the rate ratios associated with these conditions vary with age - for example the rate ratio associated with diabetes is higher at younger ages. The rate ratios of `r round(as.numeric(substr(table.agegr.all["Type 1 diabetes", 3], 1, 4)), 1)` for Type 1 diabetes and `r round(as.numeric(substr(table.agegr.all["Type 2 diabetes", 3], 1, 4)), 1)` for Type 2 diabetes are broadly similar to those reported in UK Biobank and in the OpenSAFELY studies.  We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.  An unexpected finding was that the risk associated with other forms of heart disease is higher than that associated with ischaemic heart disease.  This category includes conditions such as atrial fibrillation, cardiomyopathies and heart failure. One of the highest rate ratios is that associated with chronic kidney disease.  Prevention of nosocomial transmission in dialysis units may help to reduce this risk.  Over all age groups, `r gsub("(.*\\()(.*)\\)", "\\2", freqs.all["listed.any", 2])` of severe cases had at least one of these listed conditions. \hl{ In this dataset it is not possible to examine adequately the risk associated with neoplasms, as we cannot separately identify those who were advised to shield themselves because they had active neoplasms of lymphoid or hematopoietic tissue or were receiving treatments that affect the immune system. We hope to explore this in a separate study that will include linkage to records of shielding advice.} In patients without any listed conditions, further systematic evaluation of past hospitalisation history did not reveal a sparse set of underlying conditions; instead many diagnoses were associated with severe COVID-19.

@@

A striking finding of this study was the association of severe COVID-19 with dispensing of at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.  Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that prescription of drugs for the gastrointestinal and central nervous systems, together with nutritional supplements, contributed to this association.  Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, or to more diffuse frailty especially in older persons, causal effects of drugs or direct effects of polypharmacy on susceptibility cannot be ruled out.  These associations are explored in a separate paper.

## Relevance to policy
As lockdown restrictions are eased, there is general agreement that vulnerable individuals will require shielding, even if the restart of the epidemic can be slowed or suppressed by mass testing, contact tracing and isolation of those who test positive.  The "stratify and shield" policy option [@mckeigue_evaluation_2020], in which high-risk individuals comprising up to 15% of the population are shielded for a defined period while the epidemic is allowed to run relatively quickly in low-risk individuals until population-level immunity is attained, depends critically on informative risk discrimination.  So too does the similarly named "segment and shield " option [@bunnik_segmentation_2020] which has the opposite objective of keeping transmissions low.  Although for this preliminary study we have not used the full repertoire of machine learning methods available for this type of problem, we have shown that a model based on health records provides `r table.densities[3, 5]` bits of information for discrimination conditional on age and sex.  Adding this to the `r round(cases.Lambda.agesex, 2)` bits provided by age and sex gives a total information for discrimination of `r round(table.densities[3, 5] + cases.Lambda.agesex, 1)` bits.  \hl{We have shown elsewhere that this level of predictive performance would allow at least 80% of those at risk of severe or fatal disease to be allocated to a shielded group comprising no more than 15% of the population} [@mckeigue_evaluation_2020]

As awareness grows of how risk varies between individuals, individuals will seek information about their own level of risk.  A key implication of our results is that risk of severe or fatal disease is multifactorial.  The rate ratio of `r round(exp(10 * logistic.coeffs[3, 1]), 1)` associated with a 10-year increase in age is stronger than the rate ratios associated with common diseases such as asthma or Type 2 diabetes that are listed as conditions associated with high risk.  A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one quarter of those aged 60-74 years in the population have at least one of the listed conditions we examined.  It will also exclude many people at high risk because they have multiple risk factors each of small effect. \hl{A more meaningful way to score risk for an individual would be to use all available information to calculate a "COVID age" as the age at which the average risk for someone of the same sex in the population equates to the risk for the individual under study}. Thus the rate ratio of `r rr <- round(as.numeric(gsub(" .+", "", table.agegr.all["Type 1 diabetes", 3])), 1); rr` associated with Type 1 diabetes equates to an increase of `r round(log(rr) / logistic.coeffs[3, 1], 1)` years in COVID age.  In Scotland it is technically possible to use existing electronic health records to calculate a risk score for every individual in the population, though more work would be required to develop this as a basis for official advice and individual decisions.

@@

## Data availability
The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data. All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public

## Funding

No funding was received for this work.

@@

@@

p.wdists <- plotWdists(densities=full.densities, distlabels=c("Crude", "Adjusted")) +
    theme_grey(base_size = 18) +
    theme(legend.title=element_blank())

p.wdists
